% acknow.tex
% A Practical Introduction to Data Structures and Algorithm Analysis
% 3rd Edition: Shared between C++ and Java versions

\paragraph{Acknowledgments:}
It takes a lot of help from a lot of people to make a book.
I wish to acknowledge a few of those who helped to make this
book possible.
I apologize for the inevitable omissions.

Virginia Tech helped make this whole thing possible through sabbatical
research leave during Fall 1994, enabling me to get the project
off the ground.
My department heads during the time I have written the various
editions of this book, Dennis Kafura and Jack Carroll, provided
unwavering moral support for this project.
Mike Keenan, Lenny Heath, and Jeff Shaffer provided valuable
input on early versions of the chapters.
I also wish to thank Lenny Heath for many years of stimulating
discussions about algorithms and analysis (and how to teach both to
students).
Steve Edwards deserves special thanks for spending so much time
helping me on various redesigns of the \LangCPP\ and \LangJava\ code versions
for the second and third editions,
and many hours of discussion on the principles of program design.
Thanks to Layne Watson for his help with \Mathematica,
and to Bo Begole, Philip Isenhour, Jeff Nielsen, and Craig Struble for
much technical assistance.
Thanks to Bill McQuain, Mark Abrams and Dennis Kafura for answering
lots of silly questions about \LangCPP\ and \LangJava.

I am truly indebted to the many reviewers of the various editions of this
man\-u\-script.
For the first edition these reviewers included
J. David Bezek (University of Ev\-ans\-ville),
Douglas Campbell (Brig\-ham Young University), 
Karen Davis (University of Cincinnati),
Vijay Kumar Garg (University of Texas -- Austin),
Jim Miller (University of Kansas),
Bruce Maxim (University of Michigan -- Dearborn),
Jeff Parker (Agile Net\-works/\-Har\-vard),
Dana Richards (George Mason University),
Jack Tan (University of Hou\-s\-ton),
and Lixin Tao (Concordia University).
Without their help, this book would contain many more technical errors
and many fewer insights.

For the second edition, I wish to thank these reviewers:
Gurdip Singh (Kansas State University),
Peter Allen (Columbia University),
Robin Hill (University of Wyoming),
Norman Jacobson (University of California -- Irvine),
Ben Keller (Eastern Michigan University), and
Ken Bosworth (Idaho State University).
In addition, I wish to thank Neil Stewart and Frank J. Thesen for
their comments and ideas for improvement.

Third edition reviewers included Randall Lechlitner (University of
Houstin, Clear Lake) and Brian C. Hipp (York Technical College).
I thank them for their comments.

Prentice Hall was the original print publisher for the first and
second editions.
Without the hard work of many people there, none of this
would be possible.
Authors simply do not create printer-ready books on their own.
Foremost thanks go to Kate Hargett, Petra Rector, Laura Steele, and
Alan Apt, my editors over the years.
My production editors, Irwin Zucker for the second edition,
Kathleen Caren for the original \LangCPP\ version,
and Ed DeFelippis for the \LangJava\ version,
kept everything moving smoothly during that horrible rush at the end.
Thanks to Bill Zobrist and Bruce Gregory (I think) for getting me
into this in the first place.
Others at Prentice Hall who helped me along the way include Truly
Donovan, Linda Behrens, and Phyllis Bregman.
Thanks to Tracy Dunkelberger for her help in returning the copyright
to me, thus enabling the electronic future of this work.
I am sure I owe thanks to many others at Prentice Hall for their
help in ways that I am not even aware of.

I am thankful to Shelley Kronzek at Dover publications for her faith
in taking on the print publication of this third edition.
Much expanded, with both \LangJava\ and \LangCPP\ versions, and many
inconsistencies corrected, I am confident that this is the best
edition yet.
But none of us really knows whether students will prefer a free online
textbook or a low-cost, printed bound version.
In the end, we believe that the two formats will be mutually
supporting by offering more choices.
Production editor James Miller and design manager Marie Zaczkiewicz
have worked hard to ensure that the production is of the highest
quality.

I wish to express my appreciation to Hanan Samet for teaching me about
data structures.
I learned much of the philosophy presented here from
him as well, though he is not 
responsible for any problems with the result.
Thanks to my wife Terry, for her love and support, and to my daughters 
Irena and Kate for pleasant diversions from working too hard.
Finally, and most importantly, to all of the data structures students
over the years who have taught me what is important and what should
be skipped in a data structures course, and the many new insights
they have provided.
This book is dedicated to them.
\bigskip
{\obeylines\parskip=0pt\parindent=0pt
\hfill Cliff Shaffer
\hfill Blacksburg, Virginia
}
